---
title: "Model Planning and Building"
author: "Michael Messmer"
output:
  html_document:
    df_print: paged
---
<a href="index.html">Home</a>

## Introduction

An interesting aspect of tennis is that it's possible for the winner of the match to have won fewer total points than the loser of the match. This observation leads me to the following question: What is the best way to predict the total points won by the winner? Is it the total points won by the loser? Is comparing seeds a useful predictor? Or is it something else entirely?

## Setup

Let's start by pulling in the work we did for part 1.

```{r message=FALSE, error=FALSE, warning=FALSE, results='hide'}
include <- function(library_name){
  if( !(library_name %in% installed.packages()) )
    install.packages(library_name) 
  library(library_name, character.only=TRUE)
}
include("tidyverse")
include("knitr")
purl("deliverable1.Rmd", output = "part1.r")
source("part1.r")
```

Let's remind ourselves what our current data looks like.

```{r}
head(match)
```

Each row in our table is a single match. Specifically, any match where at least 1 player was seeded. It's important to remember that a seed value of 34 means that player was unseeded.

## Extra Data

As far as interesting stats, our previous table only contains seeds. However, our new dataset provides much more information that might be helpful in creating a model. It is from the same source and was obtained the same way as our first dataset, the info on which you can find <a href="deliverable1.html">here.</a>

## New Variables

This dataset has quite a few variables that will not be relevant for our analysis, and I will not be explaining those. A lot of the variables from our initial dataset are also contained in our new data and were defined in part 1, so for the sake of brevity I will not be defining them again here. Any variable that starts with "winner" applies to the winner of the match and has an equivalent stat for the loser of the match, and I will only be explaining from the perspective of the winner.

At this point in the analysis a better understanding of tennis concepts becomes more important, so I will explain some relevant information. A match is broken up into sets, a set is broken up into games, and a game is broken up into points. A player must win a certain number of points to win a game, a certain number of games to win a set, and a certain number of sets to win a match. The serving player gets to serve twice on each of their service points. A first serve is generally harder to return but is more likely to miss than a second serve. The player who is not serving is known as the returning player. A break point is when the returning player will win the game if they win the next point. The serving player is at an advantage, so generally speaking each player is expected to win their service games, which is why break points are significant.

Now to define our new variables:

* **winner_service_points_won**  
The amount of points won by the winning player while serving.

* **winner_service_points_total**  
The total amount of points played by the winning player while serving.

* **winner_break_points_converted**  
The amount of break points won by the winning player while returning.

* **winner_break_points_return_total**  
The total amount of break points played by the winning player while returning.

* **winner_return_points_won**  
The amount of points won by the winning player while returning.

* **winner_return_points_total**  
The total amount of points played by the winning player while returning.

* **winner_total_points_won**  
The total amount of points won by the winning player.

* **winner_total_points_total**  
The total amount of points played by both players, since both the winner and loser must play the same amount of points.

* **match_duration**  
How long the match lasted, in minutes.

## Tidying the Data

Once again, we need to import and tidy the data.

```{r results='hide', warning=FALSE}
suppressMessages(library("jsonlite"))
suppressMessages(new_data <- fromJSON("https://pkgstore.datahub.io/sports-data/atp-world-tour-tennis-data/match_stats_2017_unindexed_json/data/2bdd108bc9cefe392847ac8cf111adc0/match_stats_2017_unindexed_json.json"))
```

Let's take a look at our column names and make sure they're all suitable.

```{r}
colnames(new_data)
```

The names are all good, but much of this data is irrelevant to our analysis. Let's create a new table with only the important data

```{r}
stats <- tibble(
  "match_id"=new_data$match_id,
  "winner_service_points_won"=new_data$winner_service_points_won,
  "winner_service_points_total"=new_data$winner_service_points_total,
  "winner_break_points_converted"=new_data$winner_break_points_converted,
  "winner_break_points_return_total"=new_data$winner_break_points_return_total,
  "winner_return_points_won"=new_data$winner_return_points_won,
  "winner_return_points_total"=new_data$winner_return_points_total,
  "winner_total_points_won"=new_data$winner_total_points_won,
  "loser_service_points_won"=new_data$loser_service_points_won,
  "loser_service_points_total"=new_data$loser_service_points_total,
  "loser_break_points_converted"=new_data$loser_break_points_converted,
  "loser_break_points_return_total"=new_data$loser_break_points_return_total,
  "loser_return_points_won"=new_data$loser_return_points_won,
  "loser_return_points_total"=new_data$loser_return_points_total,
  "loser_total_points_won"=new_data$loser_total_points_won,
  "total_points"=new_data$winner_total_points_total,
  "match_duration"=new_data$match_duration
)
head(stats)
```

Now our table contains just the match id, which we will use to consolidate with our previous data, and the info that might be useful for our model.

```{r}
sum(is.na(stats))
```

There is no missing data in our new table, which is great. All of our data is encoded as appropriate types as well, so this data is good to go. Now we will consolidate our tables into one.

Let's start by filtering out unnecessary info from our first table.

```{r}
match <- select(match, c(tourney_slug, match_id, winner_seed, loser_seed))
```

Now we'll create a new table that combines our data.

```{r}
combined <- inner_join(match, stats, by = "match_id")
head(combined)
```

Before we create our model, we want to adjust the data to be more fit for linear regression. We want to predict winner_total_points_won, but most of our variables won't be proper predictors. For example, any variable that indicates a point won by the winner (winner_break_points_converted, winner_service_points_won, etc.) will be directly correlated to the total points won. The total_points_won is just a sum of all of the types of points won. Therefore, it would be much more interesting to look at ratios of points won to points total for each type, rather than just points won. We will create a new table that contains ratios for all of our relevant data. We will also add a new variable indicating whether or not the winner was the higher seed.

```{r}
ratios <- tibble(
  "tourney_slug"=combined$tourney_slug,
  "match_id"=combined$match_id,
  "winner_total_points_won"=combined$winner_total_points_won,
  "winner_seed"=combined$winner_seed,
  "winner_service_point_ratio"=combined$winner_service_points_won/combined$winner_service_points_total,
  "winner_break_point_ratio"=combined$winner_break_points_converted/combined$winner_break_points_return_total,
  "winner_return_point_ratio"=combined$winner_return_points_won/combined$winner_return_points_total,
  "winner_total_point_ratio"=combined$winner_total_points_won/combined$total_points,
  "loser_seed"=combined$loser_seed,
  "loser_service_point_ratio"=combined$loser_service_points_won/combined$loser_service_points_total,
  "loser_break_point_ratio"=combined$loser_break_points_converted/combined$loser_break_points_return_total,
  "loser_return_point_ratio"=combined$loser_return_points_won/combined$loser_return_points_total,
  "loser_total_point_ratio"=combined$loser_total_points_won/combined$total_points,
  "match_duration"=combined$match_duration,
  "higher_seed_won"=combined$winner_seed<combined$loser_seed
)
head(ratios)
```

One problem is that we have a few NaN values. Presumably these are cases where the type_point_total was equal to 0 and caused a divide by 0 error. Let's go through and change those values to 0, as that would be the proper ratio.

```{r}
ratios$winner_service_point_ratio[is.nan(ratios$winner_service_point_ratio)] <- 0
ratios$winner_break_point_ratio[is.nan(ratios$winner_break_point_ratio)] <- 0
ratios$winner_return_point_ratio[is.nan(ratios$winner_return_point_ratio)] <- 0
ratios$winner_total_point_ratio[is.nan(ratios$winner_total_point_ratio)] <- 0
ratios$loser_service_point_ratio[is.nan(ratios$loser_service_point_ratio)] <- 0
ratios$loser_break_point_ratio[is.nan(ratios$loser_break_point_ratio)] <- 0
ratios$loser_return_point_ratio[is.nan(ratios$loser_return_point_ratio)] <- 0
ratios$loser_total_point_ratio[is.nan(ratios$loser_total_point_ratio)] <- 0
head(ratios)
```

Now that we've got the data in the proper form, it's finally time to build our model.

## Building the Model

We'll start by loading the required library and setting our seed so our analysis will be reproducible.

```{r, warning=FALSE}
suppressMessages(include("caret"))
set.seed(384)
```

We will utilize cross validation by randomly assigning each row in our table to one of either a training set or testing set. The training set will be used to build our model and the testing set will be used to test the model.

```{r}
sample_selection <- createDataPartition(ratios$winner_total_points_won, p=.75, list=FALSE)
train <- ratios[sample_selection, ]
test <- ratios[-sample_selection, ]
```

Now we will build a linear model and perform multiple linear regression on our model. We will use winner_total_points_won as our dependent variable, as that is the variable we want to be able to predict. Each match has a unique match id, so it makes little sense to include it as a possible predictor. We will check all other variables.

```{r}
train_model <- lm(winner_total_points_won ~ . - match_id - tourney_slug, data=train)
summary(train_model)
```

For now we'll just look at the p-values, the Pr(>|t|) column. We're only concerned with variables that have a p-value < .05, as that indicates that they are statistically significant predictors. Let's create another model with only the significant predictors from our previous model.

```{r}
train_model2 <- lm(winner_total_points_won ~ winner_seed + winner_service_point_ratio + winner_break_point_ratio + winner_return_point_ratio + loser_seed + loser_break_point_ratio + match_duration, data=train)
summary(train_model2)
```

All of our predictors are statistically significant, so our model is as strong as it can be. Our adjusted R-squared value is .9101, which means that roughly 91% of the variability in our data can be explained by the model.

Now we will analyze and attempt to explain the relationship between each of our predictors and our dependent variable. The Estimate column is what we're interested in. The number indicates the amount that predictor value changes as our dependent variable value increases by one unit.

Winner_seed and loser_seed both increase slightly (meaning that the seed itself decreases) as winner_total_points_won increases. Generally, higher seeds are expected to win more easily than lower seeds, so higher seeds tend to need fewer points to win the match. Thus it makes sense that as the seed numbers increase, the winner tends to end up with more points.

Match_duration also increases with our dependent variable, which also makes sense. Generally speaking the longer a match goes on, the more points are played, and thus the more points the winner has the opportunity to win.

Winner_service_point_ratio and winner_return_point_ratio both strongly increase with the dependent variable, with service points being almost twice as potent. This indicates that having a higher ratio of points won to points played generally results in a higher total amount of points won. The exception to this would be incredibly close matches that last a long time where many points are played (and thus many points are won) but the ratio is much lower. However, our model seems to indicate that this is fairly rare.

A stat that is initially confusing is that winner_break_point_ratio decreases as our dependent variable increases. This seems to indicate that a player that misses more break point opportunities generally wins more points than a player that converts quickly. However, this makes sense if you consider the significance of break points and that more points won by the winner generally means that the match was closer. A player that capitalizes quickly on break points can win the match much more quickly (having to play fewer points) than a player that has many break point opportunities but few conversions.

Loser_break_point_ratio also decreases as our dependent variable increases, but this is less confusing. If the loser has many opportunities for break points but doesn't capitalize on them, this means that the winner has to win more points to win the overall match. If the loser has a very high break point conversion rate, then they either had breaks in a set where they were already ahead or didn't break enough times to win the match. It's incredibly rare for a player to have a high break point conversion and still lose the match.

## Testing the Model

```{r}
predictions <- train_model2 %>% predict(test)
R2(predictions, test$winner_total_points_won)
```

Our R2 value indicates that 91.5% of the variability in our test data can be explained by our model, which is quite high.

```{r}
RMSE(predictions, test$winner_total_points_won)
```

Our RMSE (Root Mean Squared Error) value tells us the average difference between the test values and our predicted values. This indicates that our model's predictions for winner_total_points_won are off by an average of 7.4 for our test data.

```{r}
RMSE(predictions, test$winner_total_points_won)/mean(test$winner_total_points_won)
```

Dividing the RMSE value by the mean of our dependent variable gives us the prediction error rate, which should be as small as possible. Ours is close to 0, which means that our model is quite accurate at predicting the test data.

<a href="deliverable1.html">Part 1</a>

<a href="deliverable3.html">Part 3</a>
